---
title: "Spectra in follow-up samples"
author: "Rosalie Griffin Waller, PhD"
output:
  html_document:
    df_print: paged
---

Notebook to:
a) calculate spectra in follow-up CoMMpass samples
b) calculate spectra OS linear predictor in follow-up samples
c) plot spectra OS linear predictor over time

### Setup ###
```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "/path/to/data/dir/") 

library(tidyverse)
library(data.table)
library(cowplot)

color.codes = c("#648FFF","#785EF0","#DC267F","#FE6100","#FFB000")
```

## Load data
```{r}
# Normalized and ComBat corrected expression estimates in follow-up samples (long data format)
exp_est = read.csv(file = "data/commpass_followup_only_combat_nrmd_expr_20210602.csv.gz") %>% data.table()
#head(exp_est)

# Spectra gene loadings (rotation matrix)
gweight = read.csv(file = "data/spectra_rotation_matrix-gene_centers-2021-06-01.csv.gz") %>% data.table()
#head(gweight)

# Baseline spectra standard deviation values
base_sd = read.csv(file = "data/commpass_baseline_spectra_sds_20210602.csv") %>% data.table()
#head(base_sd)

# Baseline spectra
spectra = read.csv(file = "data/commpass_spectra_baseline_os_20210601.csv") %>% data.table()

# Baseline spectra OS beta values to calculate linear predictor
beta_os = read.csv(file = "data/CoMMpass_survival_beta_table_20210604.csv") %>% data.table()

# Per-patient visit data (to get sample time-point)
per_vis = read.csv(file = "data/MMRF_CoMMpass_IA14_PER_PATIENT_VISIT.csv") %>% data.table()
```

**Un-melt expression estimates**
```{r}
cbat <- list("melt"=exp_est)
cbat.dt <- dcast(cbat$melt, sample_id ~ gene_id, value.var='adjlogcpkmed') # sample x gene
```

## Calcualte spectra in followup samples
```{r}
dim_time = cbind(cbat.dt[,"sample_id"],
                 scale(data.matrix(cbat.dt[,-"sample_id"]),
                       center=gweight$mean,scale = FALSE) %*%
                   data.matrix(gweight[,-c("gene_id","mean")]))
```

**Standardize to baseline spectra**
```{r}
tme.sd = cbind(dim_time[,"sample_id"],
               data.matrix(dim_time[,-"sample_id"]) %*% 
                 diag(1/base_sd$Sp_SD)) # divide by baseline spectra SD
colnames(tme.sd) = c("sample_id",gsub(x = colnames(dim_time[,-"sample_id"]),pattern = "PC",replacement = "Sp"))
```

## Write Spectra variables
```{r}
write.csv(tme.sd,file = "data/Spectra_Variables_CoMMpass_Followup_Samples.csv",
          row.names = FALSE)
```

## Calculate OS linear predictors in followup samples
```{r}
os_pcs = tme.sd %>% select("sample_id",beta_os[!is.na(OS)]$spectra)

os_pcs$os_lp = rowSums(data.matrix(os_pcs[,-"sample_id"]) %*% diag(beta_os[!is.na(OS)]$OS))
```

**Calculate in baseline samples**
```{r}
base_os_pcs = spectra %>% select("sample_id","ttcos","censos",beta_os[!is.na(OS)]$spectra)

base_os_pcs$os_lp = rowSums(data.matrix(base_os_pcs[,-c("sample_id","ttcos","censos")]) %*% diag(beta_os[!is.na(OS)]$OS))
```

**Generate short ID**
```{r}
base_os_pcs[,brief_id:=tstrsplit(sample_id,'_')[2]]
# merge
os_lp = rbind(base_os_pcs %>% select("sample_id","os_lp"),
              os_pcs %>% select("sample_id","os_lp"))
os_lp[,brief_id:=tstrsplit(sample_id,'_')[2]]
# annotate survival data
dat = merge(os_lp,base_os_pcs[,c("brief_id","ttcos","censos")],by = "brief_id")
```

**Find samples with multiple time-points**
```{r}
count.dt = as.data.frame(table(os_lp$brief_id))

plot.samples = droplevels(count.dt[count.dt$Freq>2,]$Var1)

plot.data = dat[brief_id%in%plot.samples] %>% data.table()
#summary(plot.data)
```

**Annotate sample time-point**
```{r}
plot.data
plot.data[,SPECTRUM_SEQ:=tstrsplit(sample_id,'_BM')[1]]

pdt2 = merge(plot.data,per_vis[,c("SPECTRUM_SEQ","VJ_INTERVAL","VISITDY")],by = "SPECTRUM_SEQ")
pdt2[,month:=tstrsplit(VJ_INTERVAL,' ')[2]]
pdt2$month[is.na(pdt2$month)] <- 0
pdt2$month = as.integer(pdt2$month)
#summary(pdt2$month)

pdt2$sample_year = pdt2$VISITDY/365.25
pdt2$ttcos_year = pdt2$ttcos/365.25
```

## Plot
```{r}
p1 = ggplot(data=pdt2,aes(x=sample_year,y=os_lp,color=brief_id)) +
        geom_line(size = 1,linetype = 1) +
        geom_point(size = 4,shape = 18) +
        ylab("OS Spectra Score") + 
        xlab("Years") + #ggtitle("Overall Survival") +
        scale_color_manual(values = c("#AA0DFE","#3283FE","#1CBE4F",
                                      "#C4451C","#FE00FA","#FEAF16",
                                      "#2ED9FF","#1CFFCE","#B10DA1",
                                      "#FBE426","#325A9B")) +
        theme_classic()
```

**Add survival point**
```{r}
pdt2_last = pdt2 %>% group_by(brief_id) %>% top_n(1,sample_year)
pdt2_last$patient = as.numeric(pdt2_last$brief_id) * as.numeric(pdt2_last$censos)
pdt2_last$patient = as.factor(pdt2_last$patient)

p2 = p1 + geom_segment(data = pdt2_last, size = 0.3,
                       aes(color = brief_id, x = sample_year, xend = ttcos_year,
                           y = os_lp, yend = os_lp),linetype = 3) +
  geom_rect(data = pdt2_last, mapping = aes(xmin = ttcos_year - 0.015,
                                            xmax = ttcos_year + 0.015,
                                            ymin =os_lp - 0.075,
                                            ymax = os_lp + 0.075,
                                            fill = patient),
            size=0.3) +
  geom_hline(yintercept = 1.0525, color = "dark gray") +
  scale_fill_manual(values = c("white","#1CBE4F","#B10DA1","#FBE426","#325A9B")) +
  theme(legend.position = "none")
p2
```

```{r}
ggsave(filename = "Figure_5_OS_over_time.pdf", plot = p2, 
       device = "pdf", width = 6.5, height = 4, units = "in")
```


